---
title: "Cyclistic Bike Case Study"
author: "Mohd sahib Mansoori"
date: "2025-03-25"
output: html_document
---

# **Executive Summary**  
  This analysis explores how annual members and casual riders use Cyclistic bikes differently to inform strategies for converting casual riders into members.
Key findings reveal that casual riders take longer rides, primarily on weekends, while members use bikes more frequently on weekdays for shorter commutes.
Recommendations include targeted weekend promotions, seasonal discounts, and digital campaigns. These insights aim to guide Cyclistic’s marketing team in boosting annual memberships and profitability.

**1.Introduction** 
  
  **1.1 Background:**
  
  Cyclistic is a Chicago-based bike-share company with a diverse fleet of 5,824 bikes across 692 stations. The company’s marketing director, Lily Moreno, seeks to increase annual memberships by understanding behavioral differences between casual riders (single/day-pass users) and annual members. The primary business question is: **How do annual members and casual riders use Cyclistic bikes differently?**  
  
  **1.2 Business Task:**
  
  The business task is to analyze Cyclistic’s historical trip data to identify usage trends between casual riders and annual members. Deliverables include:  
  - A clear statement of the business task  
- Description of data sources  
- Documentation of data cleaning steps  
- Summary of analysis  
- Supporting visualizations and key findings  
- Top three actionable recommendations  

### Business Task
How do annual members and casual riders use Cyclistic bikes differently?
  
  What insights can be leveraged to convert casual riders into members?
  
**1.3 Key Stakeholders**
 
- **Lily Moreno (Marketing Director):** Seeks data-driven insights to design conversion strategies.  
- **Cyclistic Executive Team:** Requires compelling evidence to approve marketing initiatives.  
- **Marketing Analytics Team:** Responsible for executing the analysis.  

**2. Data Sources**
  The data consists of Divvy 2019 Q1 and 2020 Q1 datasets, provided by Motivate International Inc. under a public license. 

## Data Preparation

- **Handling Missing Values:** Rows with missing `started_at` or `ended_at` were removed.  
- **Removing Duplicates:** Duplicate `ride_id` entries were deleted.  
- **New Columns:**  
  - `ride_length`: Calculated ride duration (HH:MM:SS).  
- `day_of_week`: Derived using `WEEKDAY()` to analyze usage patterns.

### Load Libraries and Data
```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(scales)
library(tidyr)
library(ggplot2)
library(conflicted)
conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")

# Load the datasets
setwd("E:\\Case Study Project\\Cyclictic Data csv")
q1_2019 <- read_csv("Divvy_Trips_2019_Q1.csv")
q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")
```
# Rename columns in 2019 data to match 2020
```{r colums Rename}
q1_2019 <- rename(q1_2019,
                  ride_id = trip_id,
                  rideable_type = bikeid,
                  started_at = start_time,
                  ended_at = end_time,
                  start_station_name = from_station_name,
                  start_station_id = from_station_id,
                  end_station_name = to_station_name,
                  end_station_id = to_station_id,
                  member_casual = usertype
)
```
# Convert ride_id and rideable_type to character
```{r Convert ride_id and rideable_type to character}
q1_2019 <- mutate(q1_2019,
                  ride_id = as.character(ride_id),
                  rideable_type = as.character(rideable_type)
) 
```
# Combine datasets
```{r Combine datasets}
all_trips <- bind_rows(q1_2019, q1_2020)
```
# Remove unnecessary columns
```{r Remove unnecessary columns}
all_trips <- select(all_trips, -c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, "tripduration"))
```
# Standardize member_casual labels
```{r Standardize member_casual labels}
all_trips <- mutate(all_trips,
                    member_casual = recode(member_casual,
                                           "Subscriber" = "member",
                                           "Customer" = "casual"
                    )
)
```
# Add date-related columns and calculate ride_length
```{r Add date-related columns and calculate ride_length}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(all_trips$date, "%m")
all_trips$day <- format(all_trips$date, "%d")
all_trips$year <- format(all_trips$date, "%Y")
all_trips$day_of_week <- format(all_trips$date, "%A")
all_trips$ride_length <- difftime(all_trips$ended_at, all_trips$started_at, units = "secs")
all_trips$ride_length <- as.numeric(all_trips$ride_length)
```
# Convert full day names to abbreviated names
```{r Convert full day names to abbreviated names}
all_trips <- all_trips %>%
  mutate(day_of_week = case_when(
    day_of_week == "Sunday" ~ "Sun",
    day_of_week == "Monday" ~ "Mon",
    day_of_week == "Tuesday" ~ "Tue",
    day_of_week == "Wednesday" ~ "Wed",
    day_of_week == "Thursday" ~ "Thu",
    day_of_week == "Friday" ~ "Fri",
    day_of_week == "Saturday" ~ "Sat",
    TRUE ~ as.character(day_of_week)
  ))
```
# Remove bad data (negative ride lengths or quality control entries)
```{r Remove bad data}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length < 0), ]
```
#Data Analysis
We perform descriptive analysis to identify differences between rider types.
Statistical Summaries

# Overall summary of ride_length
```{r summary of ride_length }
summary(all_trips_v2$ride_length)
```
# Compare ride lengths by rider type
```{r Compare ride lengths}
ride_length_stats <- all_trips_v2 %>%
  group_by(member_casual) %>%
  summarise(
    Mean_Ride_Length = mean(ride_length),
    Median_Ride_Length = median(ride_length),
    Max_Ride_Length = max(ride_length),
    Min_Ride_Length = min(ride_length)
  )
ride_length_stats
```
# Trends by Day of Week
We analyze the number of rides and average ride duration by day of the week.
```{r Trends by Day of Week}
# Calculate ride counts and average duration
analysis_summary <- all_trips_v2 %>%
  group_by(member_casual, day_of_week) %>%
  summarise(
    number_of_rides = n(),
    average_duration = mean(ride_length),
    .groups = "drop"
  ) %>%
  ungroup()

# Order days of the week
analysis_summary$day_of_week <- factor(analysis_summary$day_of_week,
                                       levels = c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat"))

# Display the summary table
analysis_summary
```
# Visualizations

create visualizations to share our findings with stakeholders.
```{r Number of Rides by Rider Type and Weekday}
ggplot(analysis_summary, aes(x = day_of_week, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge", color = "white", linewidth = 0.5) +
  scale_fill_manual(values = c("casual" = "coral", "member" = "cyan3"), name = "Rider Type") +
  scale_y_continuous(labels = scales::comma, breaks = seq(0, max(analysis_summary$number_of_rides), by = 25000)) +
  labs(title = "Number of Rides by Rider Type and Weekday",
       x = "Day of Week", y = "Number of Rides") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.title = element_text(size = 12, face = "bold"),
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.margin = margin(10, 10, 10, 10)
  )
```
Bar Graph

```{r Average Ride Duration by Rider Type and Weekday}
ggplot(analysis_summary, aes(x = day_of_week, y = average_duration, color = member_casual, group = member_casual)) +
  geom_line() +
  geom_point() +
  scale_color_manual(values = c("casual" = "coral", "member" = "cyan3"), name = "Rider Type") +
  labs(title = "Average Ride Duration by Rider Type and Weekday",
       x = "Day of Week", y = "Average Duration (seconds)") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.title = element_text(size = 12, face = "bold"),
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8),
    panel.grid.minor = element_blank(),
    plot.margin = margin(10, 10, 10, 10)
  )
```
Line Graph

# Prepare data for the heatmap (rides by hour, day, and rider type)
```{r Data rides by hour, day, and rider type}
hourly_usage <- all_trips_v2 %>%
  mutate(hour = lubridate::hour(started_at)) %>%
  group_by(member_casual, day_of_week, hour) %>%
  summarise(rides = n(), .groups = "drop")

# Order days of the week (Monday to Sunday)
hourly_usage$day_of_week <- factor(hourly_usage$day_of_week,
                                   levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))
```

# Create the heatmap

```{r Usage Patterns}
ggplot(hourly_usage, aes(x = hour, y = day_of_week, fill = rides)) +
  geom_tile(color = "white", linewidth = 0.5) +
  facet_wrap(~ member_casual, ncol = 2) +
  scale_fill_gradient(low = "lightpink", high = "darkred",
                      name = "Number of Rides",
                      breaks = seq(0, max(hourly_usage$rides, na.rm = TRUE), by = 5000)) +
  labs(title = "Usage Patterns by Hour and Day of Week",
       x = "Hour of Day (0-23)",
       y = "Day of Week") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.title = element_text(size = 12, face = "bold"),
    strip.text = element_text(size = 12, face = "bold"),
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8),
    panel.grid = element_blank(),
    plot.margin = margin(10, 10, 10, 10)
  )
```
Heatmap

# Identify top stations

```{r}
top_stations <- all_trips_v2 %>%
  group_by(member_casual, start_station_name) %>%
  summarise(number_of_rides = n(), .groups = "drop") %>%
  arrange(desc(number_of_rides)) %>%
  group_by(member_casual) %>%
  slice_head(n = 10) %>%
  ungroup()

all_top_stations <- unique(c(top_stations$start_station_name[top_stations$member_casual == "member"],
                             top_stations$start_station_name[top_stations$member_casual == "casual"]))

# Prepare heatmap data
heatmap_data <- expand.grid(start_station_name = all_top_stations, 
                            member_casual = c("member", "casual"),
                            day_of_week = c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat")) %>%
  mutate(day_of_week = as.character(day_of_week)) %>%
  left_join(all_trips_v2 %>% mutate(day_of_week = as.character(day_of_week)), 
            by = c("start_station_name", "member_casual", "day_of_week")) %>%
  group_by(start_station_name, member_casual, day_of_week) %>%
  summarise(number_of_rides = sum(!is.na(ride_id), na.rm = TRUE), .groups = "drop") %>%
  mutate(number_of_rides = coalesce(number_of_rides, 0))

# Order days and stations
heatmap_data$day_of_week <- factor(heatmap_data$day_of_week, 
                                   levels = c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat"))
station_order <- heatmap_data %>%
  group_by(start_station_name, member_casual) %>%
  summarise(total_rides = sum(number_of_rides), .groups = "drop") %>%
  arrange(member_casual, desc(total_rides)) %>%
  group_by(member_casual) %>%
  slice_head(n = 10) %>%
  ungroup() %>%
  pull(start_station_name) %>%
  unique()
heatmap_data$start_station_name <- factor(heatmap_data$start_station_name, levels = rev(station_order))
```
# Create the heatmap
```{r HeatMap}
ggplot(heatmap_data, aes(x = day_of_week, y = start_station_name, fill = number_of_rides)) +
  geom_tile(color = "white", linewidth = 0.5) +
  facet_wrap(~ member_casual, ncol = 2) +
  scale_fill_gradient(low = "lightpink", high = "darkred", limits = c(0, 800), 
                      name = "Number of Rides", breaks = seq(0, 800, by = 200)) +
  labs(title = "Heatmap of Popular Start Stations for Casual and Member Riders",
       x = "Day of Week", y = "Start Station") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    axis.title = element_text(size = 12, face = "bold"),
    strip.text = element_text(size = 12, face = "bold"),
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8),
    panel.grid = element_blank(),
    plot.margin = margin(10, 10, 10, 10)
  )
```

#**Key Findings**

- **Ride Frequency**: Members take more rides during weekdays, likely for commuting, while casual riders peak on weekends, indicating leisure use.

- **Ride Duration**: Casual riders have longer average ride durations, especially on weekends, compared to members’ shorter, more consistent rides.

- **Station Usage**: Casual riders frequent leisure-oriented stations (e.g., tourist attractions), while members use stations near business districts or transit hubs.

- **Hourly Usage Patterns**: Members show peak usage during morning (7-9 AM) and evening (4-6 PM) commute hours on weekdays, while casual riders have higher usage in the late morning to early afternoon (10 AM-2 PM) on weekends.

## Recommendations
1. **Introduce a Weekend-Only Membership**: Target casual riders with a cost-effective plan for weekend use, capitalizing on their peak usage days and times (10 AM-2 PM on Saturdays and Sundays).

2. **Offer Promotions for Longer Rides**: Provide discounts or incentives for extended trips to appeal to casual riders’ preference for longer durations, encouraging membership for cost savings.

3. **Targeted Advertising at Popular Casual Stations**: Deploy digital or physical marketing campaigns at stations popular with casual riders, especially during peak weekend hours (10 AM-2 PM), promoting annual memberships.

# **Conclusion**:

This analysis highlights distinct usage patterns between annual members and casual riders, providing a foundation for targeted marketing strategies to increase annual memberships. Future steps include validating these recommendations with a pilot program and analyzing additional data (e.g., seasonal trends) to expand insights.

# **Appendix Section**

### **Purpose of the Appendix**

- **Transparency**: This section explains how you created the R Markdown report, making your process transparent to Kaggle viewers, recruiters, or peers.

- **Educational Value**: It serves as a guide for others learning R Markdown, demonstrating best practices for reproducible reporting.

- **Portfolio Enhancement**: Including this meta-reflection shows your understanding of documentation workflows, which is a valuable skill in data analysis.

## Appendix: Building This Report in R Markdown
This section outlines the steps I followed to create this reproducible report using R Markdown, providing transparency into my documentation process and serving as a guide for others interested in creating similar reports.

### Set Up the R Markdown Environment

I began by ensuring the necessary tools were installed:
  - Installed the `rmarkdown` package using `install.packages("rmarkdown")` to enable R Markdown functionality.
- Installed required analysis packages: `tidyverse`, `scales`, and `tidyr` using `install.packages(c("tidyverse", "scales", "tidyr"))`.

- Created a new R Markdown file in RStudio via `File > New File > R Markdown`, setting the title to "Cyclistic Bike-Share Case Study" and the output format to HTML for its interactivity.

The YAML header was configured as follows to define the report's metadata:
```yaml
---
title: "Cyclistic Bike-Share Case Study"
author: "[Your Name]"
date: "`r Sys.Date()`"
output: html_document
---

## Introduction

Cyclistic is a Chicago-based bike-share company with a diverse fleet of 5,824 bikes across 692 stations.

### Business Task
Analyze Cyclistic's historical bike trip data to understand how annual members and casual riders use the service differently, aiming to design marketing strategies to convert casual riders into annual members.

### Stakeholders
- **Lily Moreno**: Director of Marketing
- **Cyclistic Marketing Analytics Team**
- **Cyclistic Executive Team**
  
  ## Data Preparation
  
  The data consists of Divvy 2019 Q1 and 2020 Q1 datasets, provided by Motivate International Inc. under a public license.

### Load Libraries and Data

{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(scales)
library(tidyr)

## Load the datasets
q1_2019 <- read_csv("Divvy_Trips_2019_Q1.csv")
q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")

### **Breakdown of Steps**

- **Structure the Report**: organized the report Data Cleaning

Removing missing values.(e.g Remove missing `started_at` or `ended_at'.)

Standardizing column names.(e.g Rename the column to data consistent)

Filtering out incorrect data entries.(e.g delete Duplicate entries)

- **Embed Code, Visualizations, and Summaries**: Descriptive Analysis and Key Trends
Ride Duration.

Calculate the mean, median, max, and min ride lengths for both user groups.

Finding: Casual riders tend to have longer ride durations, suggesting leisure usage, especially on weekends.

Rider Frequency:
Determine the number of rides by day of the week.

Finding: Annual members typically ride more frequently on weekdays, aligning with commuting patterns.

 **Render the Report**: Click "Knit" in RStudio to execute code, embed visualizations, and generate a polished HTML/PDF report, ensuring reproducibility and stakeholder-ready insights. 

**Share and Archive**:  Bar Charts: Illustrate the number of rides by day of the week segmented by user type.
Line Graphs: Depict trends in average ride duration by weekday, highlighting the contrast between casual and member usage.

Heatmap:members commuting during rush hours, casual riders using bikes for leisure mid-day on weekends